---
title: "Statistical Analysis Summary"
author: "Jim Scotland"
date: "March 11, 2018"
output: 
  html_document: 
    fig_height: 8
    fig_width: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=TRUE, include=FALSE, paged.print=FALSE}
library(xml2)
library(readxl)
library(readr)
library(tidyr)
library(dplyr)
library(purrr)
library(ggplot2)
library(directlabels)
library(lubridate)
library(broom)

# Set options to limit sci notation and decimal places
options(scipen = 999, digits = 3)

# =======================================================================
# 
# Code Needed to Load Analysis Datasets
# Complete r file for data import available here:
# Springboard_Data_Science/01_data_import.R
# 
# =======================================================================

# Data accessed at
# https://wonder.cdc.gov/

# Import CDC Suicide Data (edited version with additional footer data deleted) 
suicides_df <- read_tsv("data_edited/CDC_FirearmSuicide_1999-2016.txt")

# Remove duplicate/empty columns, make Rate numeric
suicides_df$Notes <- NULL
suicides_df$'State Code' <- NULL
suicides_df$'Year Code' <- NULL
suicides_df$`Crude Rate` <- as.numeric(suicides_df$`Crude Rate`)

# Standardize and sepcify variable names (plan to merge w/ homicide data)
suicides_df <- suicides_df %>%
  rename(state = State) %>%
  rename(year = Year) %>%
  rename(sui_cnt = Deaths) %>%
  rename(sui_pop = Population) %>%
  rename(sui_rate = 'Crude Rate')

# DC & RI too few for calculation, replace NA w/ calculation
suicides_df <- mutate(suicides_df, sui_rate = ifelse(is.na(sui_rate), round(sui_cnt / sui_pop * 100000, 1), sui_rate))

# Repeat process w/ appropriate variable adjustments for homicide data
homicides_df <- read_tsv("data_edited/CDC_FirearmHomicide_1999-2016.txt")

head(homicides_df)

homicides_df$Notes <- NULL
homicides_df$'State Code' <- NULL
homicides_df$'Year Code' <- NULL
homicides_df$`Crude Rate` <- as.numeric(homicides_df$`Crude Rate`)

homicides_df <- homicides_df %>%
  rename(state = State) %>%
  rename(year = Year) %>%
  rename(hom_cnt = Deaths) %>%
  rename(hom_pop = Population) %>%
  rename(hom_rate = 'Crude Rate')

# Same issue as suicides data - replace NA hom_rate w/ calculation
homicides_df <- mutate(homicides_df, hom_rate = ifelse(is.na(hom_rate), round(hom_cnt / hom_pop * 100000, 1), hom_rate))

# Import CDC Population Data (baseline for joining suicide/homicide data)
population_df <- read_tsv("data_edited/CDC_PopEst_1990-2016.txt")
head(population_df)

# Similar adjustments for population table
population_df$Notes <- NULL
population_df$`Yearly July 1st Estimates Code` <- NULL
population_df$`State Code` <- NULL

population_df <- population_df %>%
  rename(state = State) %>%
  rename(year = `Yearly July 1st Estimates`) %>%
  rename(pop = Population)
  
# Filter for Years applicable to available CDC data
population_df <- population_df %>%
  filter(year >= 1999)

# =======================================================================
# 
# Data Merge - CDC Suicides, CDC Homicides, CDC Population
# 
# =======================================================================

# Create standard join key variable for most common table join 
join_key <- c("state", "year")

# Join Population base table w/ homicides and suicides tables
gun_deaths_df <- left_join(population_df, homicides_df, by = join_key) %>%
  left_join(suicides_df, by = join_key) %>%
  select(-hom_pop, -sui_pop)

# Replace NA in Suicides/Homicides with Mean for respective State
# Calculate hom_rate and sui_rate to replace NA values 
gun_deaths_df <- gun_deaths_df %>%
  group_by(state) %>%
  mutate(hom_cnt = ifelse(is.na(hom_cnt), as.integer(mean(hom_cnt, na.rm = TRUE)), hom_cnt)) %>%
  mutate(sui_cnt = ifelse(is.na(sui_cnt), as.integer(mean(sui_cnt, na.rm = TRUE)), sui_cnt)) %>%
  mutate(hom_rate = ifelse(is.na(hom_rate), round(hom_cnt / pop * 100000, 1), hom_rate)) %>% 
  mutate(sui_rate = ifelse(is.na(sui_rate), round(sui_cnt / pop * 100000, 1), sui_rate))

# =======================================================================
# 
# Import Region/Subregion data to join on State for higher level analysis
# 
# =======================================================================

# Regional data information accessed at
# https://www2.census.gov/geo/docs/maps-data/maps/reg_div.txt

regions_df <- read_excel("data_edited/State_FIPS_Codes.xlsx")

# Convert code fields to integer
regions_df$fips_st <- as.integer(regions_df$fips_st)
regions_df$reg_code <- as.integer(regions_df$reg_code)
regions_df$subreg_code <- as.integer(regions_df$subreg_code)

# Create region and subregion fields w/ code+name for sorting/labeling purposes
regions_df <- regions_df %>%
  unite(region, reg_code, reg_name, sep = "-", remove = FALSE) %>%
  unite(subregion, subreg_code, subreg_name, sep = "-", remove = FALSE)

# =======================================================================
# 
# Import Boston University School of Public Health Gun Law Data
# 
# =======================================================================

# Data accessed at
# https://www.statefirearmlaws.org/table.html

state_laws_df <- read.csv("data_edited/state_gun_law_database.csv")
state_codes_df <- read_xlsx("data_edited/state_gun_laws_codebook.xlsx")

# Clean up category names in State Firearm Codes
state_codes_df <- state_codes_df %>%
  select(cat_code = `Category Code`, cat = Category, sub_cat = `Sub-Category`, var_name = `Variable Name`)

# Filter state law data for 1999-2016 period only
state_laws_df <- state_laws_df %>%
  filter(year >= 1999 & year <= 2016)

# Create simplified table with total laws for general analysis
state_laws_total_df <- state_laws_df %>%
  select(state, year, lawtotal)

# Collapse 134 individual variables in to 14 larger category groupings
laws_cat_df <- state_laws_df %>%
  mutate(deal_reg = dealer + dealerh + recordsall + recordsdealerh + recordsall + 
           reportdealer + reportdealerh + reportall + reportallh + purge + residential + 
           theft + security + inspection + liability + junkgun) %>%
  mutate(buy_reg = waiting + waitingh + permit + permith + permitlaw + fingerprint +
           training + registration + registrationh + defactoreg + defactoregh + age21handgunsale + 
           age18longgunsale + age21longgunsale + age21longgunsaled + loststolen + onepermonth) %>%
  mutate(high_risk = felony + violent + violenth + violentpartial + invcommitment + 
           invoutpatient + danger + drugmisdemeanor + alctreatment + alcoholism) %>%
  mutate(bkgrnd_chk = universal + universalh + gunshow + gunshowh + universalpermit + universalpermith + 
           backgroundpurge + threedaylimit + mentalhealth + statechecks + statechecksh) %>%
  mutate(ammo_reg = ammlicense + ammrecords + ammpermit + ammrestrict + amm18 +
           amm21h + ammbackground) %>%
  mutate(poss_reg = age21handgunpossess + age18longgunpossess + age21longgunpossess + 
           gvro + gvrolawenforcement + college + collegeconcealed + elementary + opencarryh +
           opencarryl + opencarrypermith + opencarrypermitl) %>%
  mutate(conceal_reg = permitconcealed + mayissue + showing + ccrevoke + ccbackground +
           ccbackgroundnics + ccrenewbackground) %>%
  mutate(assault_mag = assault + onefeature + assaultlist + assaultregister + assaulttransfer +
           magazine + tenroundlimit + magazinepreowned) %>%
  mutate(child_acc = lockd + lockp + lockstandards + locked + capliability + capaccess +
           capuses + capunloaded + cap18 + cap16 + cap14) %>%
  mutate(gun_traff = traffickingbackground + traffickingprohibited + traffickingprohibitedh +
           strawpurchase + strawpurchaseh + microstamp + personalized) %>%
  mutate(stnd_grnd = nosyg) %>%
  mutate(pre_empt = preemption + preemptionbroad + preemptionnarrow) %>%
  mutate(immunity_ = immunity) %>%
  mutate(dom_viol = mcdv + mcdvdating + mcdvsurrender + mcdvsurrendernoconditions +
           mcdvsurrenderdating + mcdvremovalallowed + mcdvremovalrequired + incidentremoval +
           incidentall + dvro) %>%
  select(state, year, contains("_"))

# =======================================================================
# 
# Import Giffords Law Center Gun Law Data
# 
# =======================================================================

# Data accessed at 
# http://lawcenter.giffords.org/

# Import Giffords Law Center data, compiled in CSV from website data
giff_grd_df <- read.csv("data_edited/giffords_gunlawscorecard.csv")

# Import LetterGardeConverter to translate letter to numeric grade
grd_conv_df <- read.csv("data_edited/LetterGradeConverter.csv")

# Reverse numerical ordering of death_rnk to 1 for Fewest 50 for Most
# More intuitive to have rank on both reflect greater safety
giff_grd_df <- giff_grd_df %>%
  mutate(death_rnk = 51 - death_rnk)

# Calculate numerical grade from grd_conv_df table, re-arrange table
giff_grd_df <- giff_grd_df %>%
  left_join(grd_conv_df, by = c("law_grd" = "Letter")) %>%
  select(state, year, law_grd, law_score = GPA, law_rnk, death_rnk, bkgrnd_chk) %>%
  arrange(state, year)

# Create aggreagte data table w/ avg scores and ranks by state
giff_agg_df <- giff_grd_df %>%
  group_by(state) %>%
  summarise(avg_score = round(mean(law_score), 2), avg_law_rnk = round(mean(law_rnk), 2), 
            avg_death_rnk = round(mean(death_rnk), 2), avg_bkgrnd = round(mean(bkgrnd_chk), 2))

# =======================================================================
# 
# Import CDC Suicide Data ALL METHODS
# 
# =======================================================================

# Data accessed at
# https://wonder.cdc.gov/

# Import CDC Suicide Data, ALL METHODS (edited version with additional footer data deleted) 
all_suicides_df <- read_tsv("data_edited/CDC_AllSuicides_State_1999-2016.txt")

# Remove duplicate/empty columns, make Rate numeric
all_suicides_df$Notes <- NULL
all_suicides_df$'State Code' <- NULL
all_suicides_df$'Year Code' <- NULL
all_suicides_df$`Crude Rate` <- as.numeric(all_suicides_df$`Crude Rate`)

# Standardize and sepcify variable names (plan to merge w/ homicide data)
all_suicides_df <- all_suicides_df %>%
  rename(state = State) %>%
  rename(year = Year) %>%
  rename(all_sui_cnt = Deaths) %>%
  rename(all_sui_pop = Population) %>%
  rename(all_sui_rate = 'Crude Rate')

# =======================================================================
# 
# Import Gun Ownership Data
# 
# =======================================================================

# Data accessed at
# http://injuryprevention.bmj.com/content/22/3/216

gun_own_2013_df <- read_excel("data_edited/gun_ownership_rates_2013.xlsx")


# That's all for the data importing.

```


### CDC Firearm Homicide & Suicide Data  

#### Plot total numbers of Firearm Suicides & Homicides to highlight problem at the start.  

```{r gun_deaths}
gun_deaths_df %>%
  group_by(year) %>%
  summarise(Homicide = sum(hom_cnt), Suicide = sum(sui_cnt)) %>%
  gather(key = "Firearm_Death", value = "Deaths", c(Homicide, Suicide)) %>%
  ggplot(aes(x = year, y = Deaths, color = Firearm_Death)) +
  geom_line(size = 1.5) +
  expand_limits(y = 0) +
  ylab("Annual CDC Firearm Fatality Totals") +
  xlab("Year") +
  labs(color = "Firearm Deaths") +
  labs(title = "CDC Annual Firearm Deaths, Homicide vs Suicide", 
       subtitle = "National Totals 1999-2016") +
  labs(caption = "Centers for Disease Control Data for 1999-2016") +
  theme(legend.position = "right")
```
  

##### Total firearm suicides far exceed firearm homicides and have climbed steadily since around 2006 to a total of just under 23,000.  

#### Basic statistical summary for Firearm Homicide rates for 1999-2016  
```{r stat_sum_homicides}
gun_deaths_df %>%
  ungroup(state) %>%
  summarize(N = n(), Min = min(hom_rate), Max = max(hom_rate), Avg = mean(hom_rate), 
            Median = median(hom_rate), IQR = IQR(hom_rate), SD = sd(hom_rate))
```
##### Max of 30.7 stems from exceedingly high DC rates noted on import, will filter out DC for plots to avoid skewing of y-axis.  

#### Same summary for Firearm Suicide rates  
```{r stat_sum_suicides}
gun_deaths_df %>%
  ungroup(state) %>%
  summarize(N = n(), Min = min(sui_rate), Max = max(sui_rate), Avg = mean(sui_rate), 
            Median = median(sui_rate), IQR = IQR(sui_rate), SD = sd(sui_rate))
```
##### Average firearm suicide rate is nearly 2X the average firearm homicide rate.  

#### Run correlation on Homicide vs Suicide to check for linear relationship.  
```{r cor_homicide_suicide}
gun_deaths_df %>%
  left_join(regions_df, by = "state") %>%
  filter(usps_st != "DC") %>%
  ungroup(state) %>%
  summarize(N = n(), r2 = cor(sui_rate, hom_rate)^2)
```
##### At r2 of 0.0168, close to zero relationship between homicide and suicide rates.  

#### State level Suicide Rate vs Homicide Rate colored by Region and best-fit regression line  
```{r hom_vs_sui_scatter}
gun_deaths_df %>%
  left_join(regions_df, by = "state") %>%
  filter(usps_st != "DC") %>%
  ggplot(aes(x = hom_rate, y = sui_rate, color = region, label = usps_st)) +
  geom_text() +
  stat_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(color = "Region") +
  ylab("CDC Firearm Suicide Rate") +
  xlab("CDC Firearm Homicide Rate") +
  labs(title = "Firearm Suicide Rates by Firearm Homicide Rates", 
       subtitle = "Rate: Deaths per 100,000 Population") +
  labs(caption = "Centers for Disease Control Data for 1999-2016; DC excluded to avoid skewing of scale.") +
  theme(legend.position = "right")
```
  

##### Plot confirms absence of linear relation and reveals highest homicide rates in the South (LA and MS), highest suicide rates in the West (WY, MT and AK) and large clusters of low rates for both in the Northeast (CT, RI, NY and NJ). 

#### State level "Suicide Rate - Homicide Rate" by Region
```{r state_sui-hom_rates}
gun_deaths_df %>%
  left_join(regions_df, by = "state") %>%
  filter(usps_st != "DC") %>%
  group_by(region, usps_st) %>%
  summarise(Avg_Suicide_Rate = mean(sui_rate), Avg_Homicide_Rate = mean(hom_rate),
            Suicide_Homicide_Diff = Avg_Suicide_Rate - Avg_Homicide_Rate) %>%
  arrange(desc(Suicide_Homicide_Diff), desc(usps_st)) %>%
  select(region, usps_st, Suicide_Homicide_Diff) %>%
  ggplot(aes(x = reorder(usps_st, -Suicide_Homicide_Diff), y = Suicide_Homicide_Diff, fill = region)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap(~ region, scales = "free_y") +
  labs(color = "Region") +
  ylab("Avg Firearm Suicide Rate Minus Avg Firearm Homicide Rate") +
  xlab("State") +
  labs(title = "Assessing the Greater Gun Threat: Firearm Suicides vs Firearm Homicides", 
       subtitle = "Bars Indicate Amount by which Suicide Rate Exceeds Homicide Rate") +
  labs(caption = "Centers for Disease Control Data for 1999-2016; DC excluded to avoid skewing of scale.") +
  theme(legend.position = "none")
```
  

##### Firearm homicide rates exceed firearm suicide rates in only 6 of 50 states. The largest rate differentials appear in several western states, suggesting regional influences. Turn focus to firearm suicides only.  

#### Create table distinguishing gun suicides vs other suicides
```{r echo=TRUE}
sui_method_df <- all_suicides_df %>%
  left_join(gun_deaths_df, by = join_key) %>%
  select(state, year, pop, all_cnt = all_sui_cnt, all_rate = all_sui_rate, 
         gun_cnt = sui_cnt, gun_rate = sui_rate) %>%
  mutate(gun_pct = gun_cnt/all_cnt, 
         other_cnt = all_cnt - gun_cnt, 
         other_rate = all_rate - gun_rate)
summary(sui_method_df)
```
  
##### Table reveals wildly disparate data across all categories. Investigate on subregion level per indications from plot above.  

#### Boxplot OVERALL Suicide Rates Across Subregions
```{r sui_all_subregion_box}
sui_method_df %>%
  left_join(regions_df, by = "state") %>%
  group_by(subregion, year) %>%
  summarise(all_rate = sum(all_cnt)/sum(pop) * 100000,
            gun_rate = sum(gun_cnt)/sum(pop) * 100000,
            other_rate = sum(other_cnt)/sum(pop) * 100000) %>%
  ggplot(aes(x = subregion, y = all_rate, color = subregion)) +
  geom_boxplot() +
  ylab("CDC Overall Suicide Rate") +
  xlab("Year") +
  labs(title = "Regional Suicide Rates ALL Methods", 
       subtitle = "Rate: Deaths per 100,000 Population") +
  labs(caption = "Centers for Disease Control Data for 1999-2016") +
  theme(legend.position = "right") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```
  
##### Coastal states have lower overall suicide rates, with the mountain region much higher which may be a partial result of significantly lower overall population levels.  

#### Boxplot of OVERALL State Suicide Rates Grouped by Population Quartile
```{r sui_all_state_box_popq}
sui_method_df %>%
  left_join(regions_df, by = "state") %>%
  group_by(subregion, usps_st, year) %>%
  summarise(all_rate = sum(all_cnt)/sum(pop) * 100000,
            gun_rate = sum(gun_cnt)/sum(pop) * 100000,
            other_rate = sum(other_cnt)/sum(pop) * 100000,
            avg_suicides = mean(all_cnt),
            avg_pop = mean(pop)) %>%
  ggplot(aes(x = reorder(usps_st, -all_rate), y = all_rate, fill = subregion)) +
  geom_boxplot() +
  facet_wrap(~ ntile(avg_pop, 4), scales = "free_x") +
  labs(fill = "Subregion") +
  ylab("CDC Overall Suicide Rate") +
  xlab("State") +
  labs(title = "Overall Suicide Rates by State Grouped by Population Quantile", 
       subtitle = "Rate: Deaths per 100,000 Population") +
  labs(caption = "Centers for Disease Control Data for 1999-2016") +
  theme(legend.position = "right")
```
  
##### Plot dispels issues of small Mountain states population driving higher suicide rates. Mountain states post highest overall suicide rates across all four population quartiles.  

#### Lineplot of Suicide Rates by Method Across Subregions
```{r subreg_sui_method}
sui_method_df %>%
  left_join(regions_df, by = "state") %>%
  group_by(subregion, year) %>%
  summarise(all_rate = sum(all_cnt)/sum(pop) * 100000,
            gun_rate = sum(gun_cnt)/sum(pop) * 100000,
            other_rate = sum(other_cnt)/sum(pop) * 100000) %>%
  gather(key = "Method", value = "rate", c(other_rate, gun_rate))  %>%
    ggplot(aes(x = year, y = rate, color = Method)) +
    geom_line(size = 1) +
    facet_wrap(~ subregion) +
    expand_limits(y = 0) +
    ylab("CDC Suicide Rates, Firearm & Other") +
    xlab("Year") +
    labs(title = "Regional Suicide Rates by Firearm vs Other Methods", 
         subtitle = "Rate: Deaths per 100,000 Population") +
    labs(caption = "Centers for Disease Control Data for 1999-2016") +
    theme(legend.position = "right")
```
  
##### Time series plot of suicides broken out by firearm/other indicates major regional differences. Coastal states exhibit much lower firearm rates at levels that appear more steady compared to other methods and other states.  

#### Many-Mini Plot of Suicide Rates by Method at State Level
```{r many_mini_sui_method}
sui_method_df %>%
  left_join(regions_df, by = "state") %>%
  group_by(state, year) %>%
  summarise(all_rate = sum(all_cnt)/sum(pop) * 100000,
            gun_rate = sum(gun_cnt)/sum(pop) * 100000,
            other_rate = sum(other_cnt)/sum(pop) * 100000) %>%
  gather(key = "Method", value = "rate", c(other_rate, gun_rate))  %>%
  ggplot(aes(x = year, y = rate, color = Method)) +
  geom_line(size = 1) +
  facet_wrap(~ state) +
  ylab("CDC Suicide Rates, Firearm & Other") +
  xlab("Year") +
  labs(title = "State Level Suicide Rates by Firearm vs Other Methods", 
       subtitle = "Rate: Deaths per 100,000 Population, 1999-2016") +
  labs(caption = "Centers for Disease Control Data for 1999-2016") +
  theme(legend.position = "bottom") +
  theme(axis.title.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.x=element_blank())
```
  
##### Displays variation by state with large variations in small population states jumping out.

#### Partition States into Above/Below Average Suicide Rate, Calculate Percentage Using Firearm  
```{r calc_gunpct, echo=FALSE}
sui_method_df %>%
  group_by(year) %>%
  mutate(abv_avg_rate = all_rate > mean(all_rate)) %>%
  group_by(abv_avg_rate) %>%
  summarise(n = n(), deaths = sum(all_cnt), avg_gun_pct = mean(gun_pct))
```
  
##### Since 1999, guns accounted for an average 58% of suicides in states with above average suicide rates and 48% of suicides in states with below average rates (average rates calculated annually).  

#### Plot of Percentage Suicides Using Firearm, States Partitioned into Above/Below Average Suicide Rate
```{r plot_gunpct}
sui_method_df %>%
  group_by(year) %>%
  mutate(Above_Average_Suicide_Rate = all_rate > mean(all_rate)) %>%
  group_by(year, Above_Average_Suicide_Rate) %>%
  summarise(n = n(), deaths = sum(all_cnt), avg_gun_pct = mean(gun_pct)) %>%
  ggplot(aes(x = year, y = avg_gun_pct, fill = Above_Average_Suicide_Rate)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ifelse(Above_Average_Suicide_Rate, "Above Average Suicide Rate", "Below Average Suicide Rate")) +
  ylab("Percentage of Suicides Using Firearm") +
  xlab("Year") +
  labs(title = "Percentage of Deaths by Firearm, Partition States by Above/Below Average Overall Suicide Rates", 
       subtitle = "Rate: Deaths per 100,000 Population") +
  labs(caption = "Centers for Disease Control Data for 1999-2016") +
  theme(legend.position = "none")
```

### Gun Ownership Rates  

#### Statistical Summary of National Gun Ownerhip Rates
```{r stats_nat_ownrate, echo=FALSE}
gun_own_2013_df %>%
  summarize(N = n(), Min = min(own_rate), Max = max(own_rate), Avg = mean(own_rate), 
            Median = median(own_rate), IQR = IQR(own_rate), SD = sd(own_rate))
```

##### Average ownership rate of 33% ranging from 5% to 61.7%.  

#### Summary at Regional Level
```{r stats_reg_ownrate, echo=FALSE}
gun_own_2013_df %>%
  left_join(regions_df, by = "state") %>%
  group_by(region) %>%
  summarize(N = n(), Min = min(own_rate), Max = max(own_rate), Avg = mean(own_rate), 
            Median = median(own_rate), IQR = IQR(own_rate), SD = sd(own_rate))
```

##### Below average rates in Northeast, above average in West and South - regional dynamic again.  

#### Boxplot of Ownership Rates by Region
```{r plot_reg_ownrate}
gun_own_2013_df %>%
  left_join(regions_df, by = "state") %>%
  ggplot(aes(x = region, y = own_rate, color = region, label = usps_st)) +
  geom_boxplot() +
  geom_text(position = "jitter") +
  labs(color = "Region") +
  ylab("Gun Ownership Rate") +
  xlab("Region") +
  labs(title = "Gun Ownership Rates by Region", 
       subtitle = "Household Gun Ownership Rates for 2013") +
  labs(caption = "Data cited by Kalesan B, Villarreal MD, Keyes KM, et al Gun ownership and social gun culture Injury Prevention 2016;22:216-220.") +
  theme(legend.position = "none")
```

### Gun Ownership Rates vs CDC Suicide Rates
 
#### Regression Plot FIREARM Suicide Rate (FSR) vs Gun Ownership Rate for 2013  
```{r lr_ownrate_v_fsr_all}
gun_deaths_df %>%
  filter(year == 2013) %>%
  left_join(gun_own_2013_df, by = "state") %>%
  left_join(regions_df, by = "state") %>%
  ggplot(aes(x = own_rate, y = sui_rate, label = usps_st, color = region)) +
  geom_text() +
  stat_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(color = "Region") +
  ylab("CDC Firearm Suicide Rate (2013)") +
  xlab("Gun Ownership Rate") +
  labs(title = "Firearm Suicide Rates by Gun Ownership Rates", 
       subtitle = "Household Gun Ownership Rates for 2013, CDC Rate: Deaths per 100,000 Population") +
  labs(caption = "Data cited by Kalesan B, Villarreal MD, Keyes KM, et al Gun ownership and social gun culture Injury Prevention 2016;22:216-220.") +
  theme(legend.position = "bottom")

```
  
#### Regression Calculation  
```{r lr_ownrate_fsr_all, echo=FALSE}
gun_deaths_df %>%
  filter(year == 2013) %>%
  left_join(gun_own_2013_df, by = "state") %>%
  ungroup(state) %>%                                 # Don't know why needed to ungroup, but it worked?
  summarize(N = n(), r2 = cor(own_rate, sui_rate)^2)
```
  
##### The r2 of 0.547 indicates solid relationship between suicide rate and gun ownership.  
  
  
#### Regression Plot FIREARM Suicide Rate (FSR) vs Gun Ownership Rate by Region
```{r}
gun_deaths_df %>%
  filter(year == 2013) %>%
  left_join(gun_own_2013_df, by = "state") %>%
  left_join(regions_df, by = "state") %>%
  ggplot(aes(x = own_rate, y = sui_rate, label = usps_st, color = region)) +
  geom_text() +
  stat_smooth(method = "lm", se = FALSE) +
  facet_grid(. ~ region)  +
  labs(color = "Region") +
  ylab("CDC Firearm Suicide Rate (2013)") +
  xlab("Gun Ownership Rate") +
  labs(title = "Regional Firearm Suicide Rates by Gun Ownership Rates", 
       subtitle = "Household Gun Ownership Rates for 2013, CDC Rate: Deaths per 100,000 Population") +
  labs(caption = "Data cited by Kalesan B, Villarreal MD, Keyes KM, et al Gun ownership and social gun culture Injury Prevention 2016;22:216-220.") +
  theme(legend.position = "none")
```
  
#### Regression Calculations
```{r lr_ownrate_fsr_reg echo=FALSE}
gun_deaths_df %>%
  filter(year == 2013) %>%
  left_join(gun_own_2013_df, by = "state") %>%
  left_join(regions_df, by = "state") %>%
  group_by(region) %>%
  summarize(N = n(), r2 = cor(own_rate, sui_rate)^2)
```
  
##### The r2 values range from 0.360 in Midwest to 0.487 in South. The lack of ownership data creates limitations.  
  

