---
title: "Milestone Report"
author: "Jim Scotland"
date: "Revised: April 9, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract  

This project will explore the often overlooked issue of firearm suicide in the United States. The analysis will include an overall picture of the degree and distribution of firearm suicides and an investigation of the key state-level variables of gun ownership rates and gun control legislation. Linear regression models will be developed to test the relationship between ownership levels, gun law categories and state firearm suicide rates. Finally, specific classes of gun laws will be evaulated to determine their potential effect in reducing firearm suicide deaths.  

## The Data  

The [Centers for Disease Control and Prevention](https://wonder.cdc.gov/) provides the state-level firearm homicide and suicides rates as well as overall suicide rates annually for the period of 1999 to 2016. Limited state gun law data comes from the [Giffords Law Center](http://lawcenter.giffords.org/) for 2014 to 2016. Annual gun law data for 1999 to 2016 was accessed at the Boston University School of Public Health [State Firearm Law database](https://www.statefirearmlaws.org/index.html). State level gun ownership data is exceedingly rare. This analysis was developed utilizing 2013 gun ownership survey results published in the journal [Injury Prevention](http://injuryprevention.bmj.com/content/22/3/216) for a paper by Bindu Kalesan et al., titled _Gun Ownership and Social Gun Culture_.  

## Data Issues: Gun Ownership Rates 

Gun ownership data presents conflicting issues. The survey data acquired from the Kalesan Injury Prevention article covers only the single year of 2013. This limits the ability to analyze shifts in state level gun ownership rates over time. Dr. Michael Siegel at the Boston University School of Public Health has developed a [proxy ownership metric](https://www.ncbi.nlm.nih.gov/pubmed/23956369) that has been calculated from 1980 to 2016. This data was acquired directly from Dr. Siegel and was integrated into the  analysis. However, the proxy measurement utilizes the firearm suicide rate in its calculation. As the firearm suicide rate is the dependent variable of interest in this study, this proxy measure was dropped from the analysis. Sharp discrepancies between Siegel's proxy measure and the Kalesan 2013 survey results strongly supports the need for better data on this critical variable. (insert plot of Siegel vs Kalesan)   

## Data Issues: State Firearm Laws  

The State Firearm Law database tracks 132 separate law variables over a twenty-plus year period. For  purposes of simplification, the 132 laws were grouped into the 14 larger categories outlined in the code file. Yet, the most frequent count for many of these categories remains zero. In these cases the measure is more binary than one of degree - does the state have _any_ laws within a given category? This makes comaprisons between states more challenging, especially when combined with gun ownership data for only a single year.  (plan on inserting many-mini of state law maps)

## Preliminary Findings  






```{r cars}
summary(cars)
```

